Patchwork: Incremental Computation for Abstract Interpretation

### 2019-02-05 paper.tex
Add section w/detailed explanation of Datalog strategy

%% Title information
\title[Short Title]{Patchwork: Incremental Computation for Static Analysis}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\section{Contribution: Incremental Relation Structures}

We developed an incremental data structure to represent the abstract states necessary to back our static analyses. Our representation allows for efficient incremental joins to build up the underlying relation. We use the following naming strategy...

You are convinced (*wave hand*) that the data structures in question provides a unique performance improvement for our use case.

>>>>

%% Title information
\title[Short Title]{Patchwork: Incremental Computation for Abstract Interpretation}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\section{Datalog Strategy - Limitations + Comparison}

The Datalog-style approach to static analysis uses the following broad algorithm:

\begin{enumerate}
    \item Define a set of relations
    \item Define a set of rules that refer to these relations
    \item Transform the input program into a set of 'facts', which are judgements of membership in the relations
    \item Apply the rules to generate new facts until a fixpoint is reached
\end{enumerate}

\subsection{Example}

As an example, consider the simple points-to analysis defined by the Flix authors (From Datalog to FLIX: A Declarative Language for Fixed Points on Lattices, section 2). The approach is applied as follows:

\begin{enumerate}
    \item Define a set of relations.\\
    For this problem, there are 6 relations total - 4 "input relations", which are the base facts, and 2 "derived relations", which give the information we're interested in. They are as follows: New(v, h), Assign(v, v), Store(v, f, v), Load(v, v, f) are the inputs, and VarPointsTo(v, h) and HeapPointsTo(h, f, h) are the derived relations. Here the lowercase letters are types - v for variable, h for heap, f for field.
    \newpage

    \item Define a set of rules that refer to these relations.\\
    There are four rules that define this analysis:
    \begin{enumerate}
        \item VarPointsTo(v1, h1) :- New(v1, h1)
        \item VarPointsTo(v1, h2) :- Assign(v1, v2), VarPointsTo(v2, h2)
        \item VarPointsTo(v1, h2) :- Load(v1, v2, f), VarPointsTo(v2, h1), HeapPointsTo(h1, f, h2)
        \item HeapPointsTo(h1, f, h2) :- Store(v1, f, v2), VarPointsTo(v1, h1), VarPointsTo(v2, h2)
    \end{enumerate}
    For each of the rules, the fact on the left is added to the database if all of the facts on the right are in the database.\\

    \item Transform the input program into a set of 'facts'.
    \begin{lstlisting}[language=Java]
    ClassA o1 = new ClassA(); // object A
    ClassB o2 = new ClassB(); // object B
    ClassB o3 = o2;
    o2.f = o1;
    Object r = o3.f;
    \end{lstlisting}

    This program is transformed into the following facts:\\
    New("o1", "A")\\
    New("o2", "B")\\
    Assign("o3", "o2")\\
    Store("o2", "f", "o1")\\
    Load("r", "o3", "f")\\

    The algorithm for this transformation is not given. For this simple example, it's fairly easy to see how each fact was derived from the program. Some algorithm for this transformation must be written for any analysis in this style.\\

    \item Apply the rules to generate new facts until a fixpoint is reached.\\
    This results in the following derived relations:\\
    VarPointsTo("o1", "A") - rule 1\\
    VarPointsTo("o2", "B") - rule 1\\
    VarPointsTo("o3", "B") - rule 2\\
    HeapPointsTo("B", "f", "A") - rule 4\\
    VarPointsTo("r", "A") - rule 3\\

    In the example, the final fact is what was desired.
\end{enumerate}

\subsection{Naive Incremental Computation in the Datalog World}

In the Datalog context, an edit consists of the insertion or deletion of facts to the starting database. Wlog, we consider insertions and deletions of single facts and discuss naive incremental strategies for these edits.

In the course of the Datalog engine's run, new facts are added to the database when a rule is triggered. It is easy to construct a simple 'provenance' of each fact in the database by tagging the fact with the facts that generated it. For instance, in the example from 3.1, the VarPointsTo("o3", "B") was generated via the facts Assign("o3", "o2") and VarPointsTo("o2, "B"), the latter of which was itself generated from New("o2", "B"). Assuming that in the course of the algorithm these provenances are stored, consider now an edit which consists of the deletion of the fact New("o2", "B") from the initial database. Rather than re-computing from scratch, we can start from our finished database and do the following: delete the fact New("o2", "B"), delete all facts that transitively depend on New("o2", "B"), use the remaining facts as the initial database for a new run. Crucially, rather than re-computing all the facts from scratch, using this algorithm, any computed fact which does not depend on the deleted fact does not have to be re-computed.

Now consider an edit which consists of the insertion of an arbitrary fact. This does not affect the facts in the database, which means we can do the following: simply use the finished database + the new fact as the initial database for a new run. But, what if the fact adds to a relation that is negated in one of our rules? Then the insertion of this fact is also a deletion of the corresponding negation, and the deletion should be processed first. (Datalog engines typically have restrictions on the use of negation in their rules - not discussed here in detail)

Therefore, a naive incremental algorithm for an arbitrary set of deletions and insertions $\Delta\{d_k, i_k\}$ is as follows: we have a final database $D$. For each deletion $d_k$ , delete $d_k$ and all facts that transitively depend on it from $D$, leaving us with $D'$. Then add each inserted fact to get $D' + \{i_k\}$, and do a new run starting with this database.

\subsection{IncA's Improved Approach}

The above are the simplest two incremental strategies which many Datalog engines (even those that do not advertise themselves as incremental) implement. IncA, an explicitly incremental framework for Datalog (+ lattices), adds an additional optimization in the context of rules over lattices. Assuming all functions used are \textit{monotone}, a deletion of fact $d$ and insertion of fact $i$ where $d \sqsubseteq i$ constitutes a \textit{monotonically increasing replacement} (term introduced by IncA authors). If all functions used are monotone, then it is true that any fact derived from $d$ is also derivable from $i$ (proof left to IncA folks). Therefore, rather than deleting many facts derived from $d$ and subsequently re-deriving all of them, we don't have to perform any deletions.

Therefore IncA's improved incremental algorithm is as follows. Given a final database $D$ and an arbitrary set of deletions and insertions $\Delta\{d_k, i_k\}$, first identify which $d_c$ have corresponding $i_c$ such that $d_c \sqsubseteq i_c$. Delete each of these from $\{d_k\}$ to get the strictly smaller set $\{d'_k\}$. Then run the above naive algorithm on $\Delta\{d'_k, i_k\}$

There are some additional implementation details to IncA's algorithm that let it efficiently perform these deletions/re-derivations, but those do not change the improvement that's being made in terms of incremental computation, so we elide those details here.

\subsection{Limitations}

Once the Datalog strategy is chosen, there are several inherent limitations with respect to incremental computation. We briefly outline several of them here, and gesture toward how these are overcome using the Adapton strategy.\\

Both Flix and IncA require that all functions over lattices are \textit{monotone}. This is a non-trivial restriction in the context of static analysis, as many standard analyses use non-monotonic widening functions to ensure termination. All of these are inadmissible for Flix or IncA. Adapton has no monotonicity restriction.\\

There is an important sense in which the approach taken by IncA lacks 'history' in the sense of Adapton. We illustrate via an example. Consider an arbitrary lattice with elements $a, b, c, d$ such that the following holds: $a$ and $b$ are not ordered, $a \sqcup c = d$, $b \sqcup c = d$, $c \neq d$. Assuming that $a$ and $c$ are in the initial set of facts, an analysis run will add $d$ to the database. Now assume we make the following edit: delete $a$ and insert $b$. This is not a monotonically increasing replacement, so IncA will transitively delete all facts depending on $a$, including $d$ and any facts derived from $d$. Now after adding $b$, in the incremental run, the analysis will quickly re-discover $d$. However, since $d$ was already deleted from the database, all consequents of $d$ must be completely re-computed. Because IncA cannot 'look ahead' to see that $d$ will be re-discovered from an inserted fact, work that could have been saved is lost. This is a specific consequence of the general problem that deleting all facts deriving from a deleted fact is an over-approximation that is necessary for the Datalog approach to be sound. In contrast, Adapton memoizes the calls which use $d$ in further derivations. When $a$ is deleted $d$ is stale, but as soon as $d$ is re-discovered, the any downstream computations memo-match and do not need to be re-computed.

\subsection{Advantages}

Datalog is implementation-agnostic with respect to the evaluation strategy for the engine itself. (Is this still true for IncA? I think so...)

IncA likely outperforms a more generic strategy on edits that are friendly to it (monotonically increasing replacements).
